Design and Implementation of Probabilistic Programming Language Anglican

### 2016-11-20 paper/paper.tex
in progress

Checkpoints can be uniquely named at the compilation time;
however, at runtime a checkpoint corresponding to a single code
location may occur multiple time due to recurrent invocation of
the function containing the checkpoint. Every unique occurrence
of a checkpoint must receive a different address. An addressing
scheme based on computing of stack addresses of checkpoints was
described in the context of Lightweight
Metropolis-Hastings~\cite{WSG11}.  This scheme has advantages
over the naive scheme where dynamic occurrences of checkpoints
are numbered sequentially, however it impacts the 
performance of probabilistic programs because of the computation
cost of computing the stack addresses:
\begin{enumerate}
    \item The address size is linear in stack depth.
    \item Every function call must be augmented by symbolic
        information required to compute stack addresses.
\end{enumerate}
In addition, the above scheme can still lead to inferior
correspondence between checkpoints in different traces: in
Anglican and other probabilistic languages where distributions
are first-class objects checkpoints with incompatible arguments 
can correspond to the same stack address. Consider the following
program fragment:
\begin{lstlisting}[style=default]
(let [dist (if use-gamma
               (gamma 2. 2.)
               (normal 0. 1.))
      x (sample dist)]
  (predict x))
\end{lstlisting}
The \texttt{sample} checkpoint has the same stack address in
different traces, however the random values should not be
re-used between different distributions. Further on, in some
algorithms, such as black-box variational inference, the role of 
checkpoint addresses is semantic rather than heuristic ---
appropriate correspondence must be established
between checkpoints in different traces for the algorithm to
work.

To overcome the above problems, Anglican uses a different scheme
which is almost as efficient as the scheme based on stack addresses for
re-use of previously drawn values, produces addresses of constant size,
and allows manual computation of checkpoint addresses at runtime
when the default automatic scheme is insufficient.

\begin{verbatim}
A simple heuristic gives the same acceptance rate as with interpreted Anglican on all examples,
including Neil’s Locomotion, as follows:

  a) each sample form has a unique id, generated at compile time. It does anyway, and this 
      has almost zero cost.
  b) address is a pair 
        [id-of-sample-form number-of-previous-occurences]
      the occurences are of the same form.
  c) if a sequence of random choices with the same id is interrupted, the count is padded up 
  by a small power of 2, for example 16.

For example, if random choices

(sample S1)
(sample ‘S2 …
(sample ‘S2 …
(sample ‘S1 …
(sample ‘S1 …
(sample ‘S1 …
(sample ‘S2 …

are encountered in the trace, then the addresses are:

[S1 1]
[S2 1]
[S2 2]
[S1 16]
[S1 17]
[S1 18]
[S2 16]

This way, every occurence of random choice has a unique address, but small disturbances (like
removal or addition of a single choice of a particular kind) do not derail the entire sequence.
This aligns well with recursive patterns used in DPmem, for example.

I have some thoughts about stopping early in addition to starting late, but will do it when there
are less deadlines. Right now, the code runs MUCH MUCH faster, it should take roughly 10
hours to produce the same amount of data for Neil’s Locomotion example which was produced
during 14 DAYS of the holidays. 
\end{verbatim}

>>>>

Checkpoints can be uniquely named at the compilation time;
however, at runtime a checkpoint corresponding to a single code
location may occur multiple time due to recurrent invocation of
the function containing the checkpoint. Every unique occurrence
of a checkpoint must receive a different address. An addressing
scheme based on computing of stack addresses of checkpoints was
described in the context of Lightweight
Metropolis-Hastings~\cite{WSG11}.  This scheme has advantages
over the naive scheme where dynamic occurrences of checkpoints
are numbered sequentially, however it impacts the 
performance of probabilistic programs because of the computation
cost of computing the stack addresses:
\begin{enumerate}
    \item On every function call, a component is added to the
        address, hence the address size is linear in stack depth.
    \item Every function call must be augmented by symbolic
        information required to compute stack addresses.
\end{enumerate}
In addition, the above scheme can still lead to inferior
correspondence between checkpoints in different traces: in
Anglican and other probabilistic languages where distributions
are first-class objects checkpoints with incompatible arguments 
can correspond to the same stack address. Consider the following
program fragment:
\begin{lstlisting}[style=default]
(let [dist (if use-gamma
               (gamma 2. 2.)
               (normal 0. 1.))
      x (sample dist)]
  (predict x))
\end{lstlisting}
The \texttt{sample} checkpoint has the same stack address in
different traces, however the random values should not be
re-used between different distributions. Further on, in some
algorithms, such as black-box variational inference, the role of 
checkpoint addresses is semantic rather than heuristic ---
appropriate correspondence must be established
between checkpoints in different traces for the algorithm to
work.

To overcome the above problems, Anglican uses a different scheme
which is almost as efficient as the scheme based on stack addresses for
re-use of previously drawn values, as produces addresses of constant size,
and allows manual computation of checkpoint addresses at runtime
when the default automatic scheme is insufficient. According to
the scheme:
\begin{itemize}
    \item A checkpoint may accept an auxiliary argument --- the
        checkpoint identifier. If specified, the identifier is the first
        argument of a checkpoint. For example \texttt{(sample
        'x1  (normal 0 1)} defines a \texttt{sample} checkpoint
        with identifier \texttt{x1}.
    \item If the identifier is omitted, a unique identifier is
        generated automatically as a fresh symbol.
    \item At runtime, the address of a checkpoint invocation has
        the form
        \texttt{[}\textit{checkpoint-identifier
        number-of-previous-occurrences}\texttt{]}, where the
        occurrences are of a checkpoint with the same checkpoint
        identifier.
    \item If a sequence of invocations of the same checkpoint is
        interrupted by a different checkpoint, the number of
        previous occurrences is rounded up to a multiple of
        an integer. For efficiency, a small power of 2 is used
        such as $2^4 = 16$.
\end{itemize}

For example, if checkpoint invocations
\begin{lstlisting}[style=default]
    (sample 'C1 ...)
    (sample 'C2 ...) 
    (sample 'C2 ...) 
    (sample 'C1 ...)
    (sample 'C1 ...)
    (sample 'C1 ...)
    (sample 'C2 ...) 
    (sample 'C3 ...) 
\end{lstlisting}
are encountered in the trace, then the addresses are:
\begin{verbatim}
    [C1 0]
    [C2 0]
    [C2 1]
    [C1 16]
    [C1 17]
    [C1 18]
    [C2 16]
    [C3 0]
\end{verbatim}
This way, each occurrence of a checkpoint has unique address,
but small disturbances (like removal or addition of a single
checkpoint with a particular identifier) are
unlikely to derail the entire sequence.
